On The Subspace Learning for Network Attack Detection

### 2019-05-27 docs/thesis/chapters/2_eigen_similarity/eigen_similarity.tex
prepare for thesis writing
first deliverable

\chapter{Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks}
\label{ch:2_mos_eig_sim}

@@

In the context of anomaly-based schemes, this thesis proposes a statistical approach based on signal processing techniques for the detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, this thesis models the network traffic using a signal processing formulation as a composition of three components: legitimate traffic, malicious traffic and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). To the best of our knowledge there is no similar model in the literature. The proposed technique is based on eigenvalue analysis, model order selection (MOS) and similarity analysis. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we also evaluate the accuracy and performance of the proposed framework applied to a experimental scenario and to the DARPA 1998 dataset \citep{osanaiye2016distributed}, which is a well known network traffic dataset. Furthermore, this proposed approach has its accuracy evaluation based on eigen similarity analysis for extracting detailed information about accurate time and network ports under attack.

The performed experiments show that synflood, fraggle and port scan attacks can be detected accurately and with great detail in an automatic and blind fashion, applying signal processing concepts for traffic modeling and through approaches based on MOS and eigen similarity analysis. The main contributions of the proposed framework are the capability to blindly detect time frames under network attack via MOS and eigen analysis, and the detailed identification of the network attack via eigen similarity analysis.

This chapter is organized as follows. In Section \ref{sec:2_relatedworks}, related works are discussed. Section \ref{sec:2_datamodel} presents the data model and the evaluated datasets. Section \ref{sec:2_prop_getv} describes the proposed framework for blind and automatic detection of flood and probe attacks. Section \ref{sec:2_experimentalresults} discusses the experimental validation and presents the results, and Section \ref{sec:2_Complexity} discusses the computational complexity of the proposed framework and evaluates the required processing time for tested scenarios. Section \ref{sec:2_conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{apx:a_mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Principal Component Analysis (PCA) is a statistical technique commonly used for dimensionality reduction. It uses an orthogonal transformation to convert a set of correlated variables into a set of linearly uncorrelated variables, where the first principal components have the largest variance. PCA has been used in attack detection \cite{almotairi2009technique}. However, PCA requires human intervention in order to identify abnormalities based on the eigenvalues profiles, if used without complementary techniques.

Callegari \emph{et al} \cite{Zonglin2009} propose a PCA-based method for identifying the traffic flows responsible for an anomaly detected at the aggregate level and evaluated their proposal through a dataset with synthetic anomalies added in the data. However, Callegari \emph{et al} focus on flood attack detection, not addressing probe attack detection, and their approach relies on visual analysis.

Lee \emph{et al.} \cite{Lee2013} propose OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al.}, the framework applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al.} only evaluate their proposed scheme for covariance analysis, while we adopt an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for flood and probe attacks, respectively.

Besides being prone to higher errors and false positives, such human intervention makes PCA not useful for real time applications. Therefore, in order to automate the analysis of eigenvalues profile, model order selection (MOS) schemes should be incorporated.

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work requires a training method to produce a prediction model for normal daily traffic and presents limitations on identification of behaviors without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work is not applied to probe and flood attack detection, simultaneously.

The data collected from honeypot systems, such as captured traffic and operating system logs, can be analyzed to obtain information about attack techniques, general trends of threats and exploits. Blind automatic detection of malicious traffic techniques have been developed for honeypots in \cite{david2011blind,da2012improved}. However, traffic on honeypot is simpler than real network traffic, because there are no running legitimate applications, due to the fact that honeypots emulate behavior of a host within a network to deceive and lure attackers \cite{zakaria2012review}. Since honeypots do not generate legitimate traffic, the amount of data captured in honeypots is significantly lower in comparison to a Network Intrusion Detection System (NIDS), which captures and analyzes the largest possible amount of network traffic \cite{david2011blind}. MOS for blind identification of malicious activities in honeypots was proposed by us in \cite{david2011blind}, which evaluated criteria for selecting the model order, through simulations and comparing the order of the resulting model with the true model order.

@@

Several approaches for network attack detections uses the KDD 99 \cite{ji2016multi,ahmed2016survey,osanaiye2016distributed,bhuyan2014network} datasets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 dataset are criticized by for the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled datasets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} dataset is the refined version of the KDD 99 dataset that redundant data records are removed, in order to avoid biased classifications. Additionally, some approaches uses simulated \cite{callegari2011novel} scenarios or non-public datasets their evaluations. Since the proposed approach relies on a packet level analysis and the KDD 99 and NSL-KDD datasets adopt a traffic aggregation by connections, we consider the use of a experimental scenario on a real network and the DARPA 1998 dataset, which is the source for the creation of the KDD 99 and NSL-KDD datasets. Note that the proposed approach is not based on learning or classification techniques, which are more susceptible to biased results caused by the issues in the DARPA/KDD datasets.

@@

This section presents details of the experimental scenario and the selected cases of the DARPA dataset, along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:2_DataCollection} describes the environment and scenario adopted in order to reproduce flood and probe attacks. Subsection \ref{sec:2_ModelingData} presents how network traffic can be modeled as signal superposition. Subsection \ref{sec:2_SynfloodFraggleandPortscan} details the traffic of synflood, fraggle and port scan attacks, and Subsection \ref{sec:2_DarpaDataset} discusses the use of the DARPA dataset for evaluation of the proposed approach.

@@

The environment of the analyzed scenario is composed of two computers and one router with access to the Internet and to an internal network, where the simulation of legitimate traffic, noise, flood and port scan attacks are performed. During the traffic generation, one computer assumes the role of the attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:2_fig1}.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:2_fig1}
\end{figure}

Note that the set of network traffic is modeled as legitimate, noise and malicious traffic, where the victim performs legitimate activities, that can be characterized by web access. In many organizations this type of traffic is predominant, since most of corporate services are web-based, such as: web pages, customized web-based systems and cloud services. It is possible to characterize the traffic of a DHCP service as an example of noise associated with the transport layer. For malicious traffic, three types of networks attacks are evaluated: synflood, fraggle and port scan. These attacks are reproduced using well-known security tools, such as Nmap\footnote{http://nmap.org} to port scan, Metasploit\footnote{http://www.metasploit.com} to synflood and Hping\footnote{http://hping.org} to lead the fraggle attack.

@@

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e., specific information such as origin IP or day and time of the attack are not considered. Therefore, from the entire log information, we just consider the timestamp (for sequencing), port type and port number.

@@

By modeling the dataset as a signal superposition, the network traffic (\textbf{X}) can be characterized as a mixture of three components: legitimate traffic (\textbf{U}), noise (\textbf{N}) and malicious traffic (\textbf{A}), according to the following expression:

\begin{equation}\label{eq:eq01}
	\boldsymbol{X}^{(q)} = \boldsymbol{U}^{(q)} + \boldsymbol{N}^{(q)} + \boldsymbol{A}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time grouping of network traffic. The matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns. Each row represents a communication port (TCP port or UDP port), and each column represents time bins having a appropriate size, such as one minute. Each element $x_{m,n}^{(q)}$ stands for the number of times that the port $m$ appears at the $n$-th minute, at the $q$-th time frame.

The legitimate traffic $\boldsymbol{U}^{(q)}$ is characterized by the traffic from user's operations. When a user accesses a web page, for example, there is the corresponding TCP/IP traffic to request the page, as well as there is the traffic required to domain name resolution. Figure \ref{fig:2_fig3} depicts an example of the legitimate traffic obtained during experiments.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:2_fig3}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modeled as noise $\boldsymbol{N}^{(q)}$. The automatic acquisition service of logical IP network address (DHCP) is an example of noise. Independently of any user operation, the machine receives an IP address, since it is configured to perform a DHCP address request. Figure \ref{fig:2_fig4} depicts an example of noise in a network traffic, represented by traffic to ports 67 and 68.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:2_fig4}
\end{figure}

The traffic coming from a malicious activity, such as a synflood or fraggle attack, is represented by the matrix $\boldsymbol{A}^{(q)}$. For this work we only consider the traffic from port scanning and flood attacks.

We define that if the obtained $\#\boldsymbol{A}^{(q)} ≠ 0$, then there is malicious traffic in the evaluated time frame $q$, on the other hand, if the $\#\boldsymbol{A}^{(q)} = 0$, then there is no malicious traffic. This work shows how to detect the $\#\boldsymbol{A}^{(q)}$, given only the matrix $\boldsymbol{X}^{(q)}$, in order to identify malicious network traffic.

@@

The network attacks evaluated by this work are: synflood, fraggle and port scan. The first two attacks can be qualified as flood or denial of service (DoS) attacks, while the last one can be qualified as probe or port scanning attack. DoS attempts to block access to system or network resources and is implemented by either forcing targets to be unavailable through the exploiting of system vulnerabilities, or consuming resources through large amount of network traffic, characterizing flood attacks. Probe attacks scan the networks to collect information about host, such as IP addresses, ports and services.

With respect to the synflood attacks, the attacker sends a large quantity and concurrent successive SYN requests to a target, in order to consume resources and cause a DoS. Figure \ref{fig:2_fig5} depicts an example of a synflood attack carried out in a real computer network. In an interval of ten minutes, more than 210,000 packets are sent as a synflood attack. This network traffic behavior can be considered an abnormal behavior of network traffic, especially since it is concentrated in a short period of time and presents similar outstanding traffic during the time under attack.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:2_fig5}
\end{figure}

With respect to the fraggle attack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. This attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This last part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo. Figure \ref{fig:2_fig6} depicts an example of the fraggle attack in a real computer network.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:2_fig6}
\end{figure}

More than 6,000,000 malicious packets can be counted in an interval of ten minutes, which can be considered an abnormal network traffic, especially due to the concentrated traffic in a short period of time and due to the similarity of the outstanding traffic.

@@

In TCP SYN scan, a SYN packet is sent to the destination and two types of responses may occur: SYN/ACK or RST/ACK. In the first case, the destination port is in the listening state, in the second case, the destination port is not listening. At the end of each port scanning, a RST/ACK packet is sent by the system that is performing the port scan. Therefore, a full connection or a complete three-way handshake is never established, which makes the detection of the attack sender more difficult, and requires approaches able to identify probe attacks without connection establishment. The UDP scan technique sends UDP packets to the destination port, and if it responds with a \emph{ICMP port unreachable} message, then it indicates that the scanned port is closed. On the other hand, if a message is not received, then the port is considered as open.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:2_fig7}
\end{figure}

Figure \ref{fig:2_fig7} depicts an example of the port scan attack in a real computer network. Note that the traffic is composed of two packets for each TCP port and one UDP packet to each port. The incoming and outgoing packets analysis, for each port, shows the high correlation and similarity of TCP and UDP traffic during the simulated port scan attack.

\subsection{The DARPA Dataset}
\label{sec:2_DarpaDataset}

The DARPA 1998 dataset\footnote{https://www.ll.mit.edu/ideval/data/} includes 7 weeks of sniffed traffic saved into raw TCPDUMP packet data, from inside and outside origins, with labeled attacks. The attacks in this dataset can be grouped into: denial-of-service (DoS); remote to local (R2L), which is characterized by unauthorized access from a remote machine; user to root (U2R), which is characterized by unauthorized access to local super-user privileges; and probe attack. Since the proposed approach focus on flood and probe attack, the analysis concentrates on the attacks that present behaviors similar to flood or probe attack. We observe that the most cases of DoS focus on exploit system vulnerabilities instead of on flooding attack. One example is the occurrence of a neptune attack which sends 20 only SYN packets, what is a behavior that differs of the expected flood attack behavior. Therefore, there were selected the cases that simulates several network traffic or numerous connection requests, also known as flooding attack \cite{ahmed2016survey,osanaiye2016distributed}, and the cases that scan ports sending just a few packets. From the simulated probe attacks, we select the cases that rely on TCP or UDP connections.

The data modeling follows the method described by the Subsection \ref{sec:2_ModelingData}, with time frames of 20 minutes, packet counting aggregation by minute and considering the traffic to the following ports: 20, 21, 22, 23, 25, 79, 80, 88, 107, 109, 110, 113, 115, 143, 161, 389, 443.

@@

This section describes the proposed technique to detect synflood, fraggle and port scan, according to Figure \ref{fig:2_fig80}, which represents the overview of the proposed framework for detection and identification of network attacks. In Subsection \ref{sec:2_prop_LargestEigenvaluebyTimeFrames} we present the steps for extraction of the largest eigenvalue for each $q$-th time frame. Next, in Subsection \ref{sec:2_prop_MOSSchemes}, we show how to apply the eigenvalues on the MOS scheme in order to detect the attack. In Subsection \ref{sec:2_prop_EigenvalueAnalysis}, we present the eigenvalue analysis to identify the time frames detected as under attack, and the Subsection \ref{sec:2_prop_EigenSimilarityAnalysis} describes the similarity analysis evaluated for detailed attack identification.

\begin{figure}[h!]
	\centering
     \includegraphics[width=11cm]{figures/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:2_fig80}
\end{figure}

@@

The proposed attack detection algorithm starts by the data pre-processing of a network traffic log containing IP, ports and timestamp of senders and receivers. During this step, the desired information is extracted in order to classify and count packets according to the origin and destination ports, and subsequently this information is grouped by minutes and by time frames.

@@

\begin{equation}\label{eq:eq02}
\boldsymbol{y}_{m}^{(q)} = \boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}.
\end{equation}

@@

\begin{equation}\label{eq:eq03}
\boldsymbol{\hat{R}}_{yy}^{(q)} = \frac{1}{N}\boldsymbol{Y}^{(q)}\boldsymbol{Y}^{(q)^{\rm T}}.
\end{equation}

For the detection of the port scan attack, the main components are not dominated by the variables with large variance. Moreover, the portscan traffic presents a highly correlated network traffic. In order to exploit such structure, we compute the sample covariance $\boldsymbol{\hat{R}}_{zz}^{(q)}$ whose variables have zero mean and unitary standard deviation as follows

>>>>

\chapter{Eigensim - Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks}
\label{ch:2_mos_eig_sim}

@@

This chapter proposes the Eigensim, which is an approach based on signal processing concepts applied to detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, the Eigensim models the network traffic using a signal processing formulation as a composition of three components: signal, artifact and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). The Eigensim is based on eigenvalue analysis, model order selection (MOS) and similarity analysis between eigenvectors. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we evaluate the accuracy and performance of the proposed framework applied to a experimental scenario and to the DARPA 1998 data set \citep{osanaiye2016distributed}, which is a well known network traffic data set.

The performed experiments show that synflood, fraggle and port scan attacks can be detected accurately and with great detail in an automatic and blind fashion, i.e. in an unsupervised approach, applying signal processing concepts for traffic modeling and through approaches based on MOS and eigen similarity analysis. The main contributions of the proposed framework are the capability to blindly detect time frames under network attack via MOS and eigen analysis, and the detailed identification of the network attack via eigen similarity analysis.

This chapter is organized as follows. In Section \ref{sec:2_relatedworks}, related works are discussed. Section \ref{sec:2_datamodel} presents the data model and the evaluated data sets. Section \ref{sec:2_prop_getv} describes the proposed framework for blind and automatic detection of flood and probe attacks. Section \ref{sec:2_experimentalresults} discusses the experimental validation and presents the results, and Section \ref{sec:2_Complexity} discusses the computational complexity of the proposed framework and evaluates the required processing time for tested scenarios. Section \ref{sec:2_conclusionandfutureworks} draws the conclusions and the suggestions for future work. The Appendix \ref{apx:a_mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Principal Component Analysis (PCA) is a statistical technique commonly used for dimensionality reduction. It uses an orthogonal transformation to convert a set of correlated variables into a new subspace of linearly uncorrelated variables, by means of eigenvalue value decomposition, where the first principal components have the largest variance. PCA has been used in attack detection \cite{almotairi2009technique}, considering that this technique is very sensitive to outliers and adopted for classification problems. However, PCA requires human intervention in order to identify abnormalities based on the eigenvalues profiles, if used without complementary techniques.

Callegari \emph{et al} \cite{Zonglin2009} proposed a PCA-based method for identifying the traffic flows responsible for an anomaly detected at the aggregate level and evaluated their proposal through a data set with synthetic anomalies added in the data. However, Callegari \emph{et al} focus on flood attack detection, not addressing probe attack detection, and their approach relies on visual analysis.

Lee \emph{et al.} \cite{Lee2013} presented the OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al.}, the Eigensim applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al.} only evaluate their proposed scheme for covariance analysis, while we adopt an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for flood and probe attacks, respectively.

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work requires a training method to produce a prediction model for normal daily traffic and presents limitations on identification of behaviors without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a signal processing method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work is not applied to probe and flood attack detection, simultaneously.

The data collected from honeypot systems, such as captured traffic and operating system logs, can be analyzed to obtain information about attack techniques, general trends of threats and exploits. Blind automatic detection of malicious traffic techniques have been developed for honeypots in \cite{david2011blind,da2012improved}. However, traffic on honeypot is simpler than real network traffic, because there are no running legitimate applications, i.e. background traffic, due to the fact that honeypots emulate behavior of a host within a network to deceive and lure attackers \cite{zakaria2012review}. Since honeypots do not generate legitimate traffic, the amount of data captured in honeypots is significantly lower in comparison to a Network Intrusion Detection System (NIDS), which captures and analyzes the largest possible amount of network traffic \cite{david2011blind}. MOS for blind identification of malicious activities in honeypots was proposed by us in \cite{david2011blind}, which evaluated criteria for selecting the model order, through simulations and comparing the order of the resulting model with the true model order.

@@

Several approaches for network attack detection uses the KDD 99 \cite{ji2016multi,ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by for the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets adopted by researchers \cite{osanaiye2016distributed,bhuyan2014network}.

NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. Additionally, some approaches uses simulated \cite{callegari2011novel} scenarios or non-public data sets their evaluations. Since the proposed approach relies on a packet level analysis and the KDD 99 and NSL-KDD data sets adopt a traffic aggregation by connections, we consider the use of a experimental scenario on a real network and the DARPA 1998 data set, which is the source for the creation of the KDD 99 and NSL-KDD datas ets. Note that the proposed approach is not based on learning or classification techniques, which are more susceptible to biased results caused by the issues in the DARPA/KDD data sets.

@@

This section also presents details of the experimental scenario and the selected cases of the DARPA data set, along with a description of the data set model as a signal superposition of signal, artifact and noise, that refer to legitimate, control and malicious traffic. Subsection \ref{sec:2_DataCollection} describes the environment and scenario adopted in order to reproduce flood and probe attacks. Subsection \ref{sec:2_ModelingData} presents how network traffic can be modeled as signal superposition. Subsection \ref{sec:2_SynfloodFraggleandPortscan} details the traffic of synflood, fraggle and port scan attacks, and Subsection \ref{sec:2_Darpadata set} discusses the use of the DARPA data set for evaluation of the proposed approach.

@@

The environment of the analyzed scenario is composed by two computers and one router, with access to the Internet and to an internal network. In this scenario we performed the simulation of legitimate traffic, control, flood and port scan attacks. During the traffic generation, one computer assumes the role of the attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:2.01}. The victim generate legitimate traffic by means of ordinary use of Internet resources, and control traffic is constantly generated for purposes of network administration and monitoring.

%TODO - recreate the figure and change the concepts
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:2.01}
\end{figure}

Note that the set of network traffic is modeled as legitimate, controls and malicious traffic, which refer to signal, artifact and noise, respectively. In many organizations the web based traffic is predominant, since most of corporate services are  web pages, customized web-based systems and cloud services. It is possible to characterize the traffic of a DHCP service as an example of artifact associated with the transport layer, as well as it would be possible to classify heart beat traffic and other seasonal and controlled traffic as artifact. For malicious traffic, three types of networks attacks are evaluated: synflood, fraggle and port scan. Here we refer to port scan as an attack, according to \cite{ahmed2016survey,moustafa2019holistic}, however it is one approach usually adopted for acquire information in order to perform an attack. These attacks are reproduced using well-known security tools, such as Nmap\footnote{http://nmap.org} for port scan, Metasploit\footnote{http://www.metasploit.com} for synflood and Hping\footnote{http://hping.org} to lead the fraggle attack.

@@

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e. specific information such as origin or destination IP, behavioral pattern or content of the attack are not considered. Therefore, IP spoofing or data encryption would not cause impact to the proposed approach and evaluation, since our proposal only relies on the timestamp (for sequencing) and port number.

@@

By modeling the data set as a signal superposition, the network traffic (\textbf{X}) can be characterized as a mixture of three components: legitimate traffic (\textbf{U}), artifact (\textbf{A}) and malicious traffic (\textbf{N}), according to the following expression:

\begin{equation}\label{eq:2.01}
	\boldsymbol{X}^{(q)} = \boldsymbol{U}^{(q)} + \boldsymbol{A}^{(q)} + \boldsymbol{N}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time aggregation of network traffic. The matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns, where each row represents a communication port, and each column represents time bins of a defined size, such as one minute. Each element $x_{m,n}^{(q)}$ stands for the number packets that appears at $n$-th minute for the port $m$, during the $q$-th time frame.

The legitimate traffic $\boldsymbol{U}^{(q)}$ is characterized by the traffic from user's ordinary operations, such as when performing access to web pages, as well when there is the traffic required to domain name resolution. Figure \ref{fig:2.03} depicts an example of the legitimate traffic simulated during the experiments.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:2.03}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modeled as artifact $\boldsymbol{A}^{(q)}$. The acquisition of logical IP network address (DHCP) is an example of artifact, where independently of any user operation, the machine receives an IP address, since it is configured to automatically perform a DHCP address request. Figure \ref{fig:2.04} depicts an example of artifact in a network traffic, by means of traffic to ports 67 and 68. The traffic coming from a malicious activity, i.e. port scanning, synflood or fraggle attacks, is represented by the matrix $\boldsymbol{N}^{(q)}$.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:2.04}
\end{figure}

We define that if the $\#\boldsymbol{X}^{(q)} ≠ 0$, then there is malicious traffic in the evaluated time frame $q$, on the other hand, if the $\#\boldsymbol{X}^{(q)} = 0$, then there is no malicious traffic. We show how to detect the $\#\boldsymbol{X}^{(q)}$, given only the matrix $\boldsymbol{X}^{(q)}$, in order to identify malicious traffic.

@@

The network attacks evaluated by this work are: synflood, fraggle and port scan. The first two attacks can be qualified as flood or denial of service (DoS) attacks, while the last one can be qualified as probe or port scanning attack.

A DoS is an attempt by an attacker to prevent legitimate access to websites by overwhelming the amount of available bandwidth or resources of the computer system. DoS is implemented by either forcing targets to be unavailable through the exploiting of system vulnerabilities, or consuming resources through large amount of network traffic, characterizing flood attacks.

Probe attacks scan computer and network systems to collect information about the host, such as open ports, topology, running software or version of technologies, in order to find vulnerabilities.

With respect to the synflood attacks, the attacker sends a large quantity and concurrent successive SYN requests to a target, in order to consume resources and cause a DoS. Figure \ref{fig:2.05} depicts an example of a synflood attack carried out in a real computer network. In an interval of ten minutes, more than 210,000 packets are sent as a synflood attack. This network traffic behavior can be considered an abnormal behavior of network traffic, especially since it is concentrated in a short period of time and presents similar outstanding traffic during the time under attack.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:2.05}
\end{figure}

Regarding the fraggle attack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. Figure \ref{fig:2.06} depicts an example of the fraggle attack in a real computer network.

The fraggle attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:2.06}
\end{figure}

The Figure \ref{fig:2.06} shows that more than 6,000,000 malicious packets can be counted in an interval of ten minutes, which can be considered an abnormal network traffic, especially due to the concentrated traffic in a short period of time and due to the similarity of the outstanding traffic.

@@

In TCP SYN scan, a SYN packet is sent to the destination and two types of responses may occur: SYN/ACK or RST/ACK. In the first case, the destination port is in the listening state, in the second case, the destination port is not listening. At the end of each port scanning, a RST/ACK packet is sent by the system that is performing the port scan. Therefore, a full connection or a complete three-way handshake is never established, which makes the detection of the attack sender more difficult, and requires approaches able to identify probe attacks without connection establishment.

The UDP scan technique sends UDP packets to the destination port, and if it responds with a \emph{ICMP port unreachable} message, then it indicates that the scanned port is closed. On the other hand, if a message is not received, then the port is considered as open.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:2.07}
\end{figure}

Figure \ref{fig:2.07} depicts an example of the port scan attack in a real computer network. Note that the traffic is composed of two packets for each TCP port and one UDP packet to each port. The incoming and outgoing packets analysis, for each port, shows the high correlation and similarity of TCP and UDP traffic during the simulated port scan attack.

\subsection{The DARPA Data set}
\label{sec:2_Darpadata set}

The DARPA 1998 data set\footnote{https://www.ll.mit.edu/ideval/data/} includes 7 weeks of sniffed traffic saved into raw TCPDUMP packet data, from inside and outside origins, with labeled attacks. The attacks in this data set can be grouped into: denial-of-service (DoS); remote to local (R2L), which is characterized by unauthorized access from a remote machine; user to root (U2R), which is characterized by unauthorized access to local super-user privileges; and probe attack. Since the proposed approach focus on flood and probe attack, the analysis concentrates on the attacks of the DARPA 1998 data set that present behaviors similar to flood or probe attack.

We observe that the most cases of DoS of DARPA 98 focus on exploit system vulnerabilities instead of on flooding attack. One example is the occurrence of a neptune attack which sends 20 SYN packets, what is a behavior that differs of the expected flooding attack behavior. Therefore, there were selected the cases that simulates several network traffic or numerous connection requests, also known as flooding attack \cite{ahmed2016survey,osanaiye2016distributed}, and the cases that scan ports sending just a few packets. From the simulated probe attacks, we select the cases that rely on TCP or UDP connections.

The data modeling follows the method described by the Subsection \ref{sec:2_ModelingData}, with time frames of 20 minutes, packet aggregation counting by minute and considering the traffic to the following ports: 20, 21, 22, 23, 25, 79, 80, 88, 107, 109, 110, 113, 115, 143, 161, 389, 443.

@@

This section describes the proposed technique to detect synflood, fraggle and port scan, according to the overview depicted in Figure \ref{fig:2.08}, which represents the proposed framework for detection and identification of network attacks. In Subsection \ref{sec:2_prop_LargestEigenvaluebyTimeFrames} we present the steps for extraction of the largest eigenvalue for each $q$-th time frame. Next, in Subsection \ref{sec:2_prop_MOSSchemes}, we show how to apply the eigenvalues on the MOS scheme in order to detect the attack. In Subsection \ref{sec:2_prop_EigenvalueAnalysis}, we present the eigenvalue analysis to identify the time frames detected as under attack, and the Subsection \ref{sec:2_prop_EigenSimilarityAnalysis} describes the similarity analysis evaluated for detailed attack identification.

\begin{figure}[h!]
	\centering
     \includegraphics[width=11cm]{figures/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:2.08}
\end{figure}

@@

The proposed attack detection algorithm starts by the data pre-processing of a network traffic log containing IP, ports and timestamp of senders and receivers. During this step, the desired information is extracted in order to classify and to count packets according to the origin and destination ports. Subsequently, this information is grouped by minutes and by time frames.

@@

\begin{equation}
\label{eq:2.02}
\boldsymbol{y}_{m}^{(q)} = \boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}.
\end{equation}

@@

\begin{equation}\label{eq:2.03}
\boldsymbol{\hat{R}}_{yy}^{(q)} = \frac{1}{N}\boldsymbol{Y}^{(q)}\boldsymbol{Y}^{(q)^{\rm T}}.
\end{equation}

For the detection of the port scan attack, the main components are not dominated by the variables with large variance. Moreover, the port scan traffic presents a highly correlated network traffic between the monitored ports. In order to exploit such structure, we compute the sample covariance $\boldsymbol{\hat{R}}_{zz}^{(q)}$ whose variables have zero mean and unitary standard deviation as follows

\begin{equation}
\label{eq:2.04}
\boldsymbol{z}_{m}^{(q)} = \frac{\boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}}{\boldsymbol{\sigma}_{m}^{(q)}}.
\end{equation}
